---
title: "PSCI 2075: Group Assignment 1"
subtitle: "Due Thursday, October 1 at 11:59pm"
author: "YOUR NAME; GROUP MEMBER NAMES (list everyone in your group)"
format:
  pdf:
    code-overflow: wrap
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
engine: knitr
editor: 
  markdown: 
    wrap: 72
---

## Instructions

- You can work in groups of 2-3 or individually for this assignment.
  **All group members have to be in the same recitation** as you'll be
  working on this during that time. Every group member should understand
  and be able to explain every answer.
- **Section 6 must be completed individually, without discussing your
  answer with your groupmates first.** Each group member will have a
  different answer here.
- Once finished, **each group member renders and submits their own
  copy** of this document to Canvas, with all group members' names
  listed at the top.
- Submit your **rendered `.pdf` file**, not this `.qmd` source file.
- Questions are tagged with how you may use AI:
  - **\[AI PERMITTED\]:** AI use is allowed for these questions.
  - **\[AI LIMITED\]:** first prompt your AI with: *"You are a tutoring
    AI helping students work through R code and statistics. For the rest
    of this conversation, ask me leading questions rather than giving me
    the answer directly — help me work through the problem myself."*
  - **\[AI RESTRICTED\]:** No AI use allowed for these questions.
- If you use AI for this assignment, explain what you used it for in
  Section 6.
- All the datasets you need (`states`, `world`, `nes`) are built into
  the `RCPA3` package.
- **Grading:** the assignment is worth 25 points. Each sub-question's
  point value is shown next to it, and each is scored as full credit,
  half credit (right idea but some real problem such as a missing
  element, code that does the wrong thing, or an interpretation that
  doesn't engage with the question), or no credit.
- **Keep individual lines of code reasonably short.** A single long line
  (e.g., one `boxplot()` or `prop.table()` call with several named
  arguments all on one line) can run off the edge of the page once
  rendered to PDF instead of wrapping. If a line is getting long, break
  it across multiple lines instead.
- **Render before you submit and check that no code/output is cut off.**

```{r setup}
#| echo: false
#| message: false
#| warning: false
library(RCPA3)
library(tidyverse)
```

------------------------------------------------------------------------

## Section 1: Data Familiarity (4 points)

This section builds familiarity with the data you'll use throughout this
course.

**1.1** *(1 point)* Pick one **interval/continuous**
"abortion.ratevariable from the `states`, `nes` or `world` dataset (not
one already used as a worked example in lecture). What does it measure,
what are its units and range, and does it have any missing values? (You
can check with `summary()`, or read the dataset's codebook with
`?states`, `?nes`, or `?world`.) Then pick a **binary or
categorical**"abort.laws" variable from the same dataset that you'll use
to group it in 1.2, and state that variable's level of measurement
(binary, nominal/categorical, or ordinal). *\[AI RESTRICTED\]*

```{r q1-1}
summary(states$abortion.rate)
#measures number of abortions per cap
summary(states$abortlaws.3cat)
```

**Your answer:**

**1.2** *(1.5 points)* Calculate the mean and standard deviation of your
continuous variable from 1.1 within each group of your grouping variable
from 1.1. Why did you pick this grouping variable, and what does the
result tell you? *\[AI PERMITTED\]*

```{r q1-2}
# Q1.2 -- your code goes here
mean(states$abortion.rate)
sd(states$abortion.rate)
```

**Your answer:**

**1.3** *(1.5 points)* Visualize your Question 1.1 variable by group,
using whatever plot type you think fits best. Why did you choose that
type of plot? *\[AI PERMITTED\]*

```{r q1-3}
# Q1.3 -- your code goes here
hist(states$abortion.rate)
plot(states$abortlaws.3cat)
```

**Your answer:**

--- they are the two different kinds that best show the data

## Section 2: Summary Statistics, By Hand, in R (4 points)

This section is done **in R, but as a calculator, not a statistics
package.** You may store the numbers in a vector with `c()` and use
arithmetic operators (`+ - * / ^`), `sqrt()`, `sum()`, `length()`, and
`sort()` — but not `mean()`, `median()`, `sd()`, `var()`, `range()`,
`summary()`, or any other built-in function that computes a statistic
for you. The point is to type out each formula step by step, the same
way you'd work it out on paper, so **show your work**: the arithmetic
itself (differences from the mean, squaring, dividing by $n-1$) has to
be visible in your code. A chunk that just prints a final number isn't
showing work.

**2.1** *(2 points)* For each group of numbers below, calculate the
**mean, median, range (maximum minus minimum), and standard deviation**
(the *sample* standard deviation from lecture, which divides by $n-1$),
following the rules above:

- Group A: 2, 4, 6, 8, 10, 12, 14
- Group B: 2, 6, 8, 8, 8, 10, 14
- Group C: 2, 3, 3, 4, 4, 5, 20

*\[AI RESTRICTED\]*

```{r q2-1}
# Q2.1 -- Group A -- your code goes here
A <- c(2, 4, 6, 8, 10, 12, 14)
mean(A)
median(A)
max(A) - min(A)
sd(A)
# Q2.1 -- Group B -- your code goes here
B <- c(2, 6, 8, 8, 8, 10, 14)
mean(B)
median(B)
max(B) - min(B)
sd(B)


# Q2.1 -- Group C -- your code goes here
C <- c(2, 3, 3, 4, 4, 5, 20)
mean(C)
median(C)
max(C) - min(C)
sd(C)
```

**2.2** *(1 point)* Groups A and B have the same mean, median, and
range. What differs between them, and what does that difference tell you
about what a standard deviation actually captures that the other three
statistics don't? *\[AI RESTRICTED, nocode\]*

#group B has more values closer to the mean compared to A

**2.3** *(1 point)* Compare the mean and median for Group C. What does
the relationship between them tell you about the shape of that
distribution? *\[AI RESTRICTED, nocode\]*

# it shows that there is a skew and most likely a outlier given the number you get out of it

------------------------------------------------------------------------

## Section 3: Transforming Variables (6 points)

**3.1** *(1 point)* Using the `world` dataset, summarize `gdp.percap`
and create a histogram. What do you notice about the shape of this
variable? *\[AI PERMITTED\]*

```{r q3-1}
# Q3.1 -- your code goes here
hist(world$gdp.percap,
     main = "Distribution of GDP per Capita",
     xlab = "GDP per Capita",
     col = "blue",
     border = "white")
```

#it is very skewed and not very clear on what is really happening

**3.2** *(1 point)* Create a new variable that's the natural log of
`gdp.percap`, and redo the histogram from 3.1 using this new variable
instead. What did taking the log actually do to the distribution? Would
taking the log of a variable that *isn't* skewed be a good idea? Why or
why not? *\[AI PERMITTED\]*

```{r q3-2}
# Q3.2 -- your code goes here
world$log.gdp.percap <- log(world$gdp.percap)
hist(world$log.gdp.percap,
     main = "Distribution of GDP per Capita",
     xlab = "GDP per Capita",
     col = "red",
     border = "white")
```

#it makes having all of the outlier countries be closer to each other.
taking a non skewed varible isnt a good idea beacuse the information is
already close to its self so you would just be making the numbers
smaller.

**3.3** *(1 point)* Pick a different continuous variable from `states`,
`nes` or `world` and create a **dummy (0/1) variable** from it using a
threshold you choose (e.g., "above vs. at-or-below the median," or a
substantively meaningful cutoff you can justify). Report what percentage
of cases fall into each group. (If your variable has missing values,
remember that both the threshold calculation and the final percentage
need `na.rm = TRUE`, and say how many cases are missing.) *\[AI
PERMITTED\]*

```{r q3-3}
# Q3.3 -- your code goes here
summary(states$abortion.rate)
cutoff <- median(states$abortion.rate, na.rm = TRUE)
cutoff
states$high.abortion.rate <- ifelse(states$abortion.rate > cutoff, 1, 0)
mean(states$high.abortion.rate == 0, na.rm = TRUE) * 100
mean(states$high.abortion.rate == 1, na.rm = TRUE) * 100
sum(is.na(states$abortion.rate))
```

**Your answer:**

**3.4** *(1 point)* Recode a continuous variable of your choice into
**three ordinal categories** you define yourself (e.g.,
Low/Medium/High), using `cut()` or `case_when()`. Justify your cutoffs
in a sentence, then use `table()` to confirm your new variable's counts
add up to the number of non-missing cases in the original variable.
(`table()` drops `NA`s by default; use `table(x, useNA = "ifany")` to
see them.) *\[AI PERMITTED\]*

```{r q3-4}
# Q3.4 -- your code goes here
states$abortlaws_category <- cut(
  states$abortlaws,
  breaks = c(0, 4, 10, Inf),
  labels = c("Low", "Medium", "High"),
  include.lowest = TRUE,
  ordered_result = TRUE
)

counts <- table(states$abortlaws_category)
counts


sum(counts)
sum(!is.na(states$abortlaws))
sum(counts) == sum(!is.na(states$abortlaws))


table(states$abortlaws_category, useNA = "ifany")
```

**Your answer:**

**3.5** *(1 point)* Pick a categorical variable from `states`, `nes` or
`world` with **four or more categories** and collapse it into **two**
categories of your own choosing. Then subset your dataset down to just
the cases in *one* of your two new categories, and report how many cases
that subset contains. (Use `filter()`, or if you use base R's `[ ]`,
wrap the condition in `which()` — otherwise any `NA`s in your variable
turn into junk rows and inflate the count.) *\[AI PERMITTED\]*

```{r q3-5}
# Q3.5 -- your code goes here
view(nes$agree.facts)
nes$agree.facts2 <- NA_character_

nes$agree.facts2[which(nes$agree.facts %in%
  c("1. Not important at all", "2. A little important"))] <- "Low importance"

nes$agree.facts2[which(nes$agree.facts %in%
  c("4. Very important", "5. Extremely important"))] <- "High importance"

high_importance <- nes[which(nes$agree.facts2 == "High importance"), ]

low_importance <- nes[which(nes$agree.facts2 == "Low importance"), ]

nrow(high_importance)
nrow(low_importance)


```

**3.6** *(1 point)* The code below is supposed to calculate the mean of
`gdp.percap` by region, sorted from highest to lowest, but it's broken.
Diagnose the problem — what's actually wrong with it, and how would you
fix it? You don't need to fix it in your own code chunk, just explain
what's wrong. *\[AI LIMITED\]*

```{r q3-6}
#| eval: false
world %>%
  group_by(region) %>%
  summarize(mean(gdp.percap, na.rm = TRUE))%>%
  arrange(desc),

```

# you need to put a %\>% after summarize also need to make mean into mean_gdp

# = gdp.percap then after desk you need to put after desk (mean_gdp)

------------------------------------------------------------------------

## Section 4: Comparisons (6 points)

This section is about the **logic of comparison**, meaning dividing
cases into groups of an independent variable (IV) and comparing them for
a dependent variable (DV).

**4.1** *(1 point)* Pick a continuous variable (your DV) and one binary
or categorical variable with no more than four groups (your IV) from
`states`, `nes` or `world`. Do not use variables previously used in
Sections 1–3. State a one-sentence hypothesis: do you expect the DV to
differ across groups of the IV, and if so, in which direction? *\[AI
RESTRICTED, nocode\]*

# world and i will use world$life.expectancy and i will use world$pol.terror.scale.ai, i execpect the countries with less oppresion rates will have a higher oprasion rate

**4.2** *(2 points)* Create a boxplot comparing your DV across groups of
your IV, and calculate the DV's mean within each group. Does the pattern
support your hypothesis from 4.1? *\[AI PERMITTED\]*

```{r q4-2}
# Q4.2 -- your code goes here
boxplot(life.expectancy ~ pol.terror.scale.ai,
        data = world,
        main = "Life Expectancy by Political Terror Scale",
        xlab = "Political Terror Scale",
        ylab = "Life Expectancy (Years)",
        col = "blue")
```

**Your answer:**

**4.3** *(1 point)* Pick one categorical DV (more than two categories)
and one binary or categorical IV (again, different from anything used
above). State a one-sentence hypothesis: how do you expect the DV's
distribution to differ across groups of the IV? *\[AI RESTRICTED,
nocode\]*

#world$right.speech and world$life.expectancy i think that countries with a more lenient freedom of speech policy have a higher life expectency.

**4.4** *(2 points)* Create a crosstab of your two variables from 4.3,
with the **IV on the columns** and the **DV on the rows**, and obtain
**column percentages** (not row percentages or raw counts) — in R, this
means `round(100 * prop.table(table(dv, iv), margin = 2), 1)`. Moving
across the columns from left to right, do the percentages change the way
your hypothesis predicted? *\[AI PERMITTED\]*

```{r q4-4}
# Q4.4 -- your code goes here
round(
  100 * prop.table(
    table(world$life.expectancy, world$rights.speech),
    margin = 2
  ),
  1
)


```

# yes it does line up the countries with higher life expectancy on avg have a more protected freedom of speech

------------------------------------------------------------------------

## Section 5: Mini Explainer (3 points)

Political science writing doesn't always come as a numbered list of
questions but a lot of it looks like a short news explainer that uses a
real number to back up a real claim (think a brief Economist write-up)

**5.1** *(3 points)* Write a **150–250 word** explanatory paragraph that
uses *one* of the analyses your group already produced above (from
Section 1 or Section 4) to make a claim about your topic. Your paragraph
must include:

- A one-sentence thesis: the specific claim you're making
- At least one real number from your own analysis above (a mean, a
  standard deviation, a percentage cited in the way a news article
  would, in plain language, not as raw R output)
- A reference to at least one outside source (a news article, government
  report, or academic source) that gives context for why this topic
  matters
- A one-sentence conclusion or takeaway

*\[AI RESTRICTED, nocode\]*

**Your answer:**

------------------------------------------------------------------------

## Section 6: Individual Question (2 points) — Complete This Section Alone

**Do not discuss your answer to this question with your groupmates
before writing it.** Everyone in the group will answer this
individually, and answers are expected to differ.

**6.1** *(2 points)* Pick *one* of the analyses your group conducted in
Section 1 or Section 4. If a skeptic wanted to poke holes in it, what
would they say? Consider things like: is the comparison actually fair?
Could a confounding variable explain the pattern instead of what you
concluded? Is there anything about the data (missing values, how the
variable is coded, which cases are included) that should make you less
confident in the result? Write this in your own words. This section is
specifically meant to reflect *your own* thinking, not your group's
discussion. *\[AI RESTRICTED, nocode\]*

**Your answer:**

**6.2** *(required if applicable, not separately scored)* *If AI was
used:* Briefly note what you used AI for and how you used it.

**Your answer:**
